Accelerating parameter inference with graphics processing units

### 2019-03-25 paper/referee/response-v0.tex

\item * Sec 3b introduces many new variables, a handful of which are
undefined. Please make sure all variables have been defined. Better
still, if possible, I believe the readability of this section could be
improved if all of the variables their definitions are placed into a
table.

\noindent \textbf{Response}:

\item * In table 1: why does the setup time reduce when more modes are
included? (comparing the first two rows of this table)

\noindent \textbf{Response}:  
With only 20 setups (waveform generation plus overlaps) per profiling run,  we anticpate that $\tau_{setup}$ is among the
least-well-constrained pieces of information in our tables.  Moreover, we perform the vast majority of our timing estimates on individual runs on busy nodes and clusters on which we
have shared access, not
dedicated hardware, so we can dump (and analyze) detailed profiling output.  We have seen small timing discrepancies from run to run for the setup due to resource
availability; for example, we also find setup time decreased when going from $\pm 2$ to $\pm 2, \pm 1$ for GPU (b).
More broadly, the CPU and GPU(b) tests were performed on the same machine and reflect the same hardware and logic
path, so almost all the variation seen in the $\tau_{setup}$ reflects variability in timing on a single machine.

As this aspect now has a more significant impact on our performance,  we will delve much more deeply into waveform generation and overlap timing in the near future in later work, as we
optimize these aspects of the code.

### 2019-04-06 paper/paper.tex
referee: basic

\begin{figure*}
% BASED ON
%   CIT: test_workflow_batch_gpu_lowlatency
% COMMAND
% ../plot_last_iterations_with.sh --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':'
% ../plot_last_iterations_with.sh --parameter mc --parameter delta_mc   --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':'
%
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_m2.png}
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_converge.png}
\caption{\label{fig:BBH:MultiIterate}\textbf{Convergence of BBH analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic binary black hole.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
  value of the underlying marginalized likelihood evaluations.   
%The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
\emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  \AddedResponse{the array $\ln{\cal L}_{\rm marg}(\bm{\lambda}_j)$  for
$j=1,2,\ldots N_{\rm eval}$ indexing all candidate sets of intrinsic parameters $\bm{\lambda}_j$ performed in that iteration},  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: The estimated evidence $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget, Monte Carlo error is not shown.
}
\end{figure*}

@@

\begin{figure*}
% COMMAND
% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_corner_mc_delta_mc.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
\caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
  value of the underlying marginalized likelihood evaluations.  
% The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
\emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, Monte Carlo error is not shown.
}
\end{figure*}

@@

\begin{figure*}
% COMMAND
% plot_posterior_corner.py  --posterior-file posterior_samples-5.dat  --posterior-file posterior_samples-4.dat   --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc  --parameter chi_eff --mc-range  '[1.19655,1.197]' --chi-max 0.025 --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None  --posterior-label 5 --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle ':'  --posterior-color 'k' --posterior-color 'r' --posterior-color 'g' --posterior-color 'b' --posterior-color 'k' --lnL-cut 50
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_corner_mc_delta_xi.png}
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_mc_cum.png}
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_meanVar.png}
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_converge.png}
\caption{\label{fig:BNS:Spin}\textbf{Demonstration of low-latency spinning analysis}:
Results for marginal posterior distributions
  of our fiducial synthetic neutron star, assuming the binary has nonprecessing spins.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
  value of the underlying marginalized likelihood evaluations.  Only evaluations with marginalized log-likelihoods within $30$ of the maximum are shown, to increase contrast. The solid curves show an analysis with SEOBNRv4\_ROM.
%  the dotted curves show an analysis with SpinTaylorT4.  \editremark{currenly only v4ROM shown} 
 \emph{Left panel } Posterior distribution
  over  $\mc$,
  $\delta=(m_1-m_2)/M$, and $\chi_{\rm eff}$.    \emph{Right panel}: Marginal 1d CDFs of $\delta$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, Monte Carlo error is not shown.
}
\end{figure*}

@@

\AddedResponse{In this work, we have focused exclusively on a simplified pipeline  to produce posterior distributions
  for detector-frame intrinsic parameters.   The  code  also produces
reliable extrinsic parameter distributions \cite{gwastro-PE-AlternativeArchitectures}.  With some postprocessing,
naturally provides joint posterior distributions for all intrinsic and extrinsic parameter distributions together;
examples of these distributions have been published \cite{LIGO-O2-Catalog}.  
Presently, rather than harvest extrinsic information from every iteration, we harvest joint intrinsic and
extrinsic information with a single final iteration, which we will implement shortly in our production pipeline.
}

\begin{acknowledgements}
DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
JL is supported by NSF PHY-1707965.  
RO'S is supported by NSF PHY-1707965 and PHY-1607520.
Y-LLF was supported in part by BNL LDRD projects No.\,17-029 and No.\,19-002, 
and New York State Urban Development Corporation, d/b/a Empire State Development, under contract No.\,AA289.
We thank the Computational Science Initiative at BNL for hosting GPU Hackathon 2018, during which a part of this work was conducted. We also thank Jeff Layton at NVIDIA for bringing CuPy to our attention.
\end{acknowledgements}

>>>>

\begin{figure*}
% BASED ON
%   CIT: test_workflow_batch_gpu_lowlatency
% COMMAND
% ../plot_last_iterations_with.sh --parameter m1 --parameter m2  --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':'
% ../plot_last_iterations_with.sh --parameter mc --parameter delta_mc   --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc.xml.gz  --composite-file all.net --quantiles None  --use-legend --posterior-color 'b' --posterior-color 'k' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':' --posterior-linestyle ':'
%
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_m2.png}
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_m1_cum.png}
% python plot_mean_variance.py --convergence-file 20190203-bbh-zerospin-batch_gpu_lowlatency_meanVar.dat  --output bbh_zerospin_lnL_meanVar
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_meanVar.png}
% python plot_convergence.py --convergence-file 20190203-bbh-zerospin-batch_gpu_lowlatency.dat
\includegraphics[width=0.45\textwidth]{figures/bbh_zerospin_lnL_converge.png}
\caption{\label{fig:BBH:MultiIterate}\textbf{Convergence of BBH analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic binary black hole.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
  value of the underlying marginalized likelihood evaluations.   
%The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
\emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  \AddedResponse{the array $\ln{\cal L}_{\rm marg}(\bm{\lambda}_j)$  for
$j=1,2,\ldots N_{\rm eval}$ indexing all candidate sets of intrinsic parameters $\bm{\lambda}_j$ performed in that iteration},  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: The estimated evidence $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget, Monte Carlo error is not shown.
}
\end{figure*}

@@

\begin{figure*}
% COMMAND
% plot_posterior_corner.py --posterior-file posterior_samples-4.dat --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc --mc-range  '[1.19655,1.19682]' --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_corner_mc_delta_mc.png}
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_mc_cum.png}
% python plot_mean_variance.py --convergence-file 20190201-bns-zerospin-nobatch_meanVar.dat  --output bns_zerospin_lnL_meanVar
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_meanVar.png}
% python plot_convergence.py --convergence-file 20190201-bns-zerospin-nobatch.dat --output bns_zerospin_lnL_converge
\includegraphics[width=0.45\textwidth]{figures/bns_zerospin_lnL_converge.png}
\caption{\label{fig:BNS:MultiIterate}\textbf{Convergence of BNS analysis: Zero spin}: Results for marginal posterior distributions
  of our fiducial synthetic neutron star.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
  value of the underlying marginalized likelihood evaluations.  
% The corresponding dotted curves show an analysis using   only the $m=\pm 2$ modes \editremark{perform both} 
\emph{Left panel } Posterior distribution
  over  $\mc$ and
  $\delta=(m_1-m_2)/M$.    \emph{Right panel}: Marginal 1d CDFs of $\mc$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, Monte Carlo error is not shown.
}
\end{figure*}

@@

\begin{figure*}
% COMMAND
% plot_posterior_corner.py  --posterior-file posterior_samples-5.dat  --posterior-file posterior_samples-4.dat   --posterior-file posterior_samples-3.dat --posterior-file posterior_samples-2.dat --posterior-file posterior_samples-1.dat --parameter mc --parameter delta_mc  --parameter chi_eff --mc-range  '[1.19655,1.197]' --chi-max 0.025 --plot-1d-extra --ci-list  '[0.9]' --truth-file /home/oshaughn/unixhome/Projects/LIGO-ILE-Applications/ILE-GPU/ILE-GPU-Paper/demos/mdc_bns.xml.gz  --composite-file all.net --quantiles None  --posterior-label 5 --posterior-label 4 --posterior-label 3 --posterior-label 2 --posterior-label 1 --use-legend --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle '-' --posterior-linestyle ':'  --posterior-color 'k' --posterior-color 'r' --posterior-color 'g' --posterior-color 'b' --posterior-color 'k' --lnL-cut 50
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_corner_mc_delta_xi.png}
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_mc_cum.png}
% python plot_mean_variance.py --convergence-file 20190203-bns-withspin-batch_gpu_lowlatency_noHM_v4ROM_meanVar.dat  --output bns_withspin_lnL_meanVar
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_meanVar.png}
% python plot_convergence.py --convergence-file 20190203-bns-withspin-batch_gpu_lowlatency_noHM_v4ROM.dat  --output bns_withspin_lnL_converge
\includegraphics[width=0.45\textwidth]{figures/bns_withspin_lnL_converge.png}
\caption{\label{fig:BNS:Spin}\textbf{Demonstration of low-latency spinning analysis}:
Results for marginal posterior distributions
  of our fiducial synthetic neutron star, assuming the binary has nonprecessing spins.  Solid contours show credible intervals; solid one-dimensional distributions
  show marginal CDFs and PDFs for the corresponding variable; and colored points indicate the location $\bm{\lambda}$ and
  value of the underlying marginalized likelihood evaluations.  Only evaluations with marginalized log-likelihoods within $30$ of the maximum are shown, to increase contrast. The solid curves show an analysis with SEOBNRv4\_ROM.
%  the dotted curves show an analysis with SpinTaylorT4.  \editremark{currenly only v4ROM shown} 
 \emph{Left panel } Posterior distribution
  over  $\mc$,
  $\delta=(m_1-m_2)/M$, and $\chi_{\rm eff}$.    \emph{Right panel}: Marginal 1d CDFs of $\delta$, showing convergence.
\emph{Bottom left}: Mean and variance of  $\ln{\cal L}_{\rm marg}$ on the evaluation points,  showing that after the
first iteration the
candidate points are consistent with the posterior (i.e., no proposed point has very low $\ln {\cal L}_{\rm marg}$).
\emph{Bottom right panel}: $Z = \int d\bm{\lambda} {\cal L}_{\rm marg}$ versus iteration number.  As systematic fitting error dominates our
error budget early on, Monte Carlo error is not shown.
}
\end{figure*}

@@

\AddedResponse{In this work, we have focused exclusively on profiling a simplified pipeline  to produce posterior distributions
  for detector-frame intrinsic parameters.   The  code  also produces
reliable extrinsic parameter distributions \cite{gwastro-PE-AlternativeArchitectures}.  With some postprocessing,
this pipeline  provides joint posterior distributions for all intrinsic and extrinsic parameter distributions together;
examples of these distributions have been published elsewhere \cite{LIGO-O2-Catalog}.  
Presently, rather than harvest extrinsic information from every iteration, we harvest joint intrinsic and
extrinsic information with a single final iteration, which we will implement shortly in our production pipeline.
}

\begin{acknowledgements}
We thank our anonymous referee for helpful feedback.
DW thanks the RIT COS and CCRG for support, and  NSF-1707965.
JL is supported by NSF PHY-1707965.  
RO'S is supported by NSF PHY-1707965 and PHY-1607520.
Y-LLF was supported in part by BNL LDRD projects No.\,17-029 and No.\,19-002, 
and New York State Urban Development Corporation, d/b/a Empire State Development, under contract No.\,AA289.
We thank the Computational Science Initiative at BNL for hosting GPU Hackathon 2018, during which a part of this work was conducted. We also thank Jeff Layton at NVIDIA for bringing CuPy to our attention.
\end{acknowledgements}
